disaggregation: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling

### 2019-12-04 article.Rnw
Formatting figures. Captions and figure references. Other minor changes

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\usepackage{amsmath}

%
<<kintroptions, echo=FALSE>>=
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  fig.width = 6,
  fig.height = 4
)
@
%

@@

For the disaggregation model we model the rate at pixel level, with the likelihood for then observed data then given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

In order to know which pixels (i.e. which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function). Additionally, an INLA mesh is built to use for the spatial field (\code{build_mesh} function). The \code{mesh.args} argument allows you to supply a list of INLA mesh parameters to control the mesh used for the spatial field. The mesh can take a while to construct, so if you want to prepare the data without buidling the mesh, the user can set the \code{makeMesh} flag to FALSE. However, you cannot then fit the disaggregation model without the mesh.

@@

We can see a summary of the data, using the generic \code{summary} function, and \code{plot} the data. The summary function returns information on how many pixels and polygons the data contains, how many pixels in the smallest and largest polygons and a summary of the covariate data. The \code{plot} functions plots a map of the polygon response data, the covariate rasters and the INLA mesh.

@@

%
<<plotdata1, echo=FALSE, dev='png', fig.width = 8, fig.height = 6>>=
data_plots$polygon
@
%

%
<<plotdata2, echo=FALSE, dev='png', fig.width = 8, fig.height = 5>>=
data_plots$covariates
@
%

%
<<plotdata3, echo=FALSE, dev='png', fig.width = 5, fig.height = 5>>=
INLA::plot.inla.mesh(dis_data$mesh, main = 'INLA mesh for spatial field')
@
%

@@

Now we can fit the disaggregation model using \code{fit_model}. Here we use a Poisson likelihood for the incident count data with a log link function. Options for the likelihood are Gaussian, Poisson and binomial, and options for the link function are logit, log and identity. The spatial field and iid effect are components of the model by default, they can be turned off using the \code{field} and \code{iid} flags. We specify all of the priors for the model in a single list. Hyperpriors for the field are given as penalised complexity priors - specify $\rho_{min}$ and $\rho_{prob}$ for the range of the field, where $P(\rho < \rho_{min}) = \rho_{prob}$, and $\sigma_{min}$ and $\sigma_{prob}$ for the variation of the field, where $P(\sigma > \sigma_{min}) = \sigma_{prob}$. Similarly, you specify penalised complexity priors for the iid effect.
The \code{iterations} argument specifies the maximum number of iterations the model can run for to find an optimal point. In order to print more verbose output you can set the \code{silent} argument to FALSE.

@@

We can get a summary and plot of the model output. The \code{summary} function gives the estimate and standard error of the fixed effect parameters in the model, as well as the negative log likelihood and in-sample performance metrics (root mean squared error, mean absolute error, pearson correlation coefficient and spearman rank correlation coefficient). The \code{plot} function produces two plots: one of the fixed effects parameters and one of the observed data against in-sample predictions.

@@

%
<<plotmodel, fig.width = 10, fig.height = 6>>=
plot(fitted_model)
@
%

@@

The \code{plot} function can be used on both the \code{predictions} and \code{uncertainty} objects.

%
<<plotmean, dev = 'png', fig.width = 10, fig.height = 7>>=
plot(model_predictions$mean_predictions)
@
%

%
<<plotuncertainty, dev = 'png', fig.width = 10, fig.height = 5>>=
plot(model_predictions$uncertainty_predictions)
@
%

Using three simple functions we have been able to fit a Bayesian spatial disaggregation model and predict pixel-level incidence rate across Madagascar using aggregated incidence data and pixel-level environmental covariates.

\section{Comparison with Markov chain Monte Carlo (MCMC)} \label{sec:comparison}

@@

%
\begin{figure}[t!]
\centering
\includegraphics[width=\linewidth]{figures/mcmc_long_trace.pdf}
\caption{Trace of the fixed effects parameters for MCMC using NUTS sampling, running the algorithm for 94 hours.}
\label{fig:mcmclong}
\end{figure}
%

@@

Future work will entail extending the \pkg{disaggregation} package to include spatio-temporal disaggregation models. This would require a spatio-temporal field as well as dynamic covariates. However, this would be significantly more computationally intensive. Additionally, tools for cross-validation could be included within the package. Cross-validation of spatial models is non-trivial due to the spatial autocorrelation in the data.

>>>>

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\usepackage{amsmath}
\usepackage{subfig}
\usepackage{float}

%
<<kintroptions, echo=FALSE>>=
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = 'H'
)
@
%

@@

For the disaggregation model we model the rate at pixel level, with the likelihood for then observed data then given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific iid effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

In order to know which pixels (i.e. which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function). Additionally, an INLA mesh is built to use for the spatial field (\code{build_mesh} function). The \code{mesh.args} argument allows the user to supply a list of INLA mesh parameters to control the mesh used for the spatial field. The mesh can take a while to construct, so to prepare the data without buidling the mesh the user can set the \code{makeMesh} flag to FALSE. However, it would then not be possible to fit the disaggregation model without the mesh.

@@

We can see a summary of the data, using the generic \code{summary} function, and \code{plot} the data. The summary function returns information on how many pixels and polygons the data contains, how many pixels in the smallest and largest polygons and a summary of the covariate data. The \code{plot} functions plots a map of the polygon response data, the covariate rasters and the INLA mesh, as shown in Figure~\ref{fig:subplotdata}.

@@

%
<<subplotdata, fig.cap='Maps of Madagascar showing the data used in the disaggregation model. These plots are produced when calling the plot function on the disag.data object.', fig.subcap=c('Number of malaria cases in each administrative unit', 'Maps of the four covariates used in the model', 'inla.mesh object that will be used to make the spatial field'), out.width='0.5\\linewidth', echo=FALSE, fig.ncol = 2, dev = 'png', fig.width = 4, fig.height = 5, fig.pos = 'H'>>=
data_plots$polygon
data_plots$covariates
INLA::plot.inla.mesh(dis_data$mesh, main = 'INLA mesh for spatial field')
@
%

@@

Now we can fit the disaggregation model using \code{fit_model}. Here we use a Poisson likelihood for the incident count data with a log link function. Options for the likelihood are Gaussian, Poisson and binomial, and options for the link function are logit, log and identity. The spatial field and iid effect are components of the model by default, they can be turned off using the \code{field} and \code{iid} flags. We specify all of the priors for the model in a single list. Hyperpriors for the field are given as penalised complexity priors - specify $\rho_{min}$ and $\rho_{prob}$ for the range of the field, where $P(\rho < \rho_{min}) = \rho_{prob}$, and $\sigma_{min}$ and $\sigma_{prob}$ for the variation of the field, where $P(\sigma > \sigma_{min}) = \sigma_{prob}$. Similarly, the user specifies penalised complexity priors for the iid effect.
The \code{iterations} argument specifies the maximum number of iterations the model can run for to find an optimal point. In order to print more verbose output the user can set the \code{silent} argument to FALSE.

@@

We can get a summary and plot of the model output. The \code{summary} function gives the estimate and standard error of the fixed effect parameters in the model, as well as the negative log likelihood and in-sample performance metrics (root mean squared error, mean absolute error, pearson correlation coefficient and spearman rank correlation coefficient). The \code{plot} function produces two plots: one of the fixed effects parameters and one of the observed data against in-sample predictions, as shown in Figure~\ref{fig:fitmodel}.

@@

\begin{figure}[!h]
\centering
%
<<plotmodel, fig.width = 10, fig.height = 6>>=
plot(fitted_model)
@
%
\caption{\label{fig:fitmodel} Plot summarising the results of the fitted model. These plots are produced when calling the \code{plot} function on the \code{fit.result} object. The fixed effects plot (left) shows the fitted parameter values with uncertainty estimation for all the fixed effects in the model. The in-sample performance plot (right) shows the predicted incidence rate values for each polygon in the data against the observed values for that polygon in the data.}
\end{figure}

@@

The \code{plot} function can be used on both the \code{predictions} and \code{uncertainty} objects, as shown in Figures~\ref{fig:subpredictmean} and \ref{fig:predictuncertainty} respectively.

%
<<plotpredictions, eval=FALSE>>=
plot(model_predictions$mean_predictions)
@
%

%
<<hiddenplotmean, echo=FALSE, results='hide', fig.show='hide'>>=
mean_predictions <- plot(model_predictions$mean_predictions)
@
%

%
<<subpredictmean, fig.cap='Maps of Madagascar showing the fine-scale predictions from the disaggregation model. These maps are produced when calling the plot function on the predictions object. The maps show the (a) mean prediction of malaria incidence rate, (b) covariate contribution to the linear predictor, and (c) field contribution to the linear predictor', fig.subcap=c('', '', ''), out.width='0.33\\linewidth', echo=FALSE, dev = 'png', fig.width = 4, fig.height = 6, fig.pos = 'H'>>=
mean_predictions$mean
mean_predictions$covariates
mean_predictions$field
@
%

\begin{figure}
\centering
%
<<plotuncertainty, dev = 'png', fig.width = 10, fig.height = 5>>=
plot(model_predictions$uncertainty_predictions)
@
%
\caption{\label{fig:predictuncertainty} Maps of Madagascar showing the fine-scale predictions of lower (2.5\%) and upper (97.5\%) credible intervals of the malaria incidence rate from the disaggregation model. These maps are produced when calling the \code{plot} function on the \code{uncertainty} object.}
\end{figure}

Using three simple functions within the \pkg{disaggregation} package we have been able to fit a Bayesian spatial disaggregation model and predict pixel-level incidence rate across Madagascar using aggregated incidence data and pixel-level environmental covariates.

\newpage
\section{Comparison with Markov chain Monte Carlo (MCMC)} \label{sec:comparison}

@@

%
\begin{figure}[t!]
\centering
\includegraphics[width=\linewidth]{figures/mcmc_long_trace.pdf}
\caption{Trace of the fixed effects parameters for MCMC using NUTS sampling, running the algorithm for 94 hours. It was run for 8000 iterations with 2000 of those as warmup.}
\label{fig:mcmclong}
\end{figure}
%

@@

Future work could be done extending the \pkg{disaggregation} package to include spatio-temporal disaggregation models. This would require a spatio-temporal field as well as dynamic covariates, and would be significantly more computationally intensive. Additionally, tools for cross-validation could be included within the package. Cross-validation of spatial models is non-trivial due to the spatial autocorrelation in the data.
